Space-time adaptivity for parabolic evolution equations

### 2020-05-26 Latex/paper.tex
minor edits in the preconditioning section

\subsubsection{A preconditioner for bilinear forms on $Y \times Y$}
Let $\Psi$ be a Riesz basis for $L_2(I)$, any $y \in Y$ is of the form
$\sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$.
Using that $\|w\|_Y \eqsim \sum_{\lambda \in \vee_\Psi} \|w_\lambda\|_V^2 < \infty$,
on $Y \times Y$, we define the bounded, symmetric, coercive bilinear form
\[
  (D_Y \sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_{\Phi}} \psi_\mu \otimes v_\mu ) := \sum_{\lambda \in \vee_\Psi} \la w_\lambda, v_\lambda \ra_V.
\]
With $E_Y^\delta$ the trivial embedding $Y^\delta \to Y$,
we set $D_Y^\delta := {E_Y^\delta}' D_Y E_Y^\delta \in \Lis(Y^\delta, {Y^\delta}')$,
whose norm and norm of its inverse are uniformly bounded in $\delta \in \Delta$.
Equipping $Y^\delta$ with a basis of the form $\cup_\lambda \psi_\lambda \otimes \Xi_\lambda^\delta$,
the matrix representation of $D_Y^\delta$ reads
\[
  \mathbf D_Y^\delta = \blockdiag [\mathbf A_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Xi_\lambda^\delta, \Xi_\lambda^\delta \ra_V.
\]
For given $\mathbf B_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta)^{-1}$, we have
\[
  (\mathbf D_Y^\delta)^{-1} \eqsim \blockdiag[\mathbf B_\lambda^\delta]_\lambda.
\]
In \S\ref{sec:mgprecond} below, we construct such $\mathbf B_\lambda^\delta$ in the form of a
\emph{multigrid preconditioner}, which can be applied in linear complexity. A
simpler approach, although at superlinear complexity, would be to perform direct
solves in the spatial axis, i.e.~$\mathbf B_\lambda^\delta := (\mathbf A_\lambda^\delta)^{-1}$.

\subsubsection{A preconditioner for bilinear forms on $X \times X$}
Let $\Sigma$ be a Riesz basis for $L_2(I)$ such that
$\{2^{-|\lambda|}\sigma_\lambda : \lambda \in \vee_\Sigma\}$ is a Riesz basis for $H^1(I)$.
Using that any $w \in X$ is of the form $\sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$,
on $X \times X$ we define the bilinear form
\[
  (D_X \sum_{\lambda \in \vee_{\Sigma}} \sigma_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_{\Sigma}} \sigma_\mu \otimes v_\mu) := \sum_{\lambda \in \vee_\Sigma} \la w_\lambda, v_\lambda \ra_V + 4^{|\lambda|} \la w_\lambda, v_\lambda \ra_{V'}.
\]
\begin{lemma}
  The symmetric bilinear form $D_X$ is bounded and coercive.
\end{lemma}
\begin{proof}
  The result follows directly from the norm equivalence
  \[
    \|w\|_X^2 \eqsim \sum_{\lambda \in \vee_\Sigma} \|w_\lambda\|_V^2 + 4^{|\lambda|} \|w_\lambda\|_{V'}^2.
  \]
  \jw{TODO prove; hier kwamen we vorige keer niet netjes uit.}
\end{proof}
Denoting with $E_X^\delta$ the trivial embedding $X^\delta \to X$, we again set
$D_X^\delta := {E_X^\delta}' D_X E_X^\delta \in \Lis(X^\delta, {X^\delta}')$, whose
norm and norm of its inverse are bounded uniformly in $\delta \in \Delta$. Equipping
$X^\delta$ with a basis of the form $\cup_\lambda \sigma_\lambda \otimes \Phi_\lambda^\delta$,
the matrix representation of $D_X^\delta$ reads
\[
  \mathbf D_X^\delta = \blockdiag[\mathbf A_\lambda^\delta + 4^{|\lambda|} \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'}]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_V.
\]
\begin{theorem}
If the collection $\mathcal O := \{W_\lambda^\delta = \tilde W_\lambda^\delta : \lambda \in \vee_\Sigma^\delta\}$
satisfies the conditions of Proposition~\ref{prop:space-infsup} with uniform inf-sup constant $\gamma>0$,
then
  \[
    (\mathbf D_X^\delta)^{-1} \eqsim \blockdiag[\mathbf C_\lambda^\delta \mathbf A_\lambda^\delta \mathbf C_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1} \quad \text{and} \quad \mathbf M_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra.
  \]
  In other words, constructing the $\mathbf C_\lambda^\delta$ leads to an optimal preconditioner on $X \times X$.
\end{theorem}
\begin{proof}

  With $\mathbf u$ the representation of $u$ in $\Phi_\lambda^\delta$, we have
  $\|u\|_{{Y^\delta}'} = \|(\mathbf A_\lambda^\delta)^{-1/2} \mathbf M_\lambda^\delta \mathbf u\|$.
For $u \in W_\lambda^\delta$, we have $\gamma \|u\|_{V'} \leq \|u\|_{(W_\lambda^\delta)'} \leq \|u\|_{V'}$, so that
  \[
    \gamma^2 \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'} \leq \mathbf M_\lambda^\delta (\mathbf A_\lambda^\delta)^{-1} \mathbf M_\lambda^\delta \leq \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'}.
  \]
  From \cite[Thm.~4]{Pearson2012}, we find the equivalence
  \[
\tfrac12\big( {\bf A}_\lambda^\delta + 4^{|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda\big)
\leq ({\bf A}_\lambda^\delta + 2^{|\lambda|}{\bf M}^\delta_\lambda) ({\bf A}_\lambda^\delta)^{-1} ({\bf A}_\lambda^\delta + 2^{|\lambda|}{\bf M}^\delta_\lambda)
\leq {\bf A}_\lambda^\delta + 4^{|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda.
\]
Therefore, given $\mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$ uniformly in $\delta$ and $|\lambda|$, the result holds.
\end{proof}
We will momentarily construct such $\mathbf C_\lambda^\delta$ as a multigrid
preconditioner. A (suboptimal) alternative is to take
$\mathbf C_\lambda^\delta := (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$
directly.

\subsubsection{A multigrid preconditioner in space}
\label{sec:mgprecond}
\jw{ray: kan jij hier iets over schrijven?}

>>>>

\subsubsection{A preconditioner for bilinear forms on $Y \times Y$}
Let $\Psi$ be a Riesz basis for $L_2(I)$, any $y \in Y$ is of the form
$\sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$.
Using that $\|w\|_Y \eqsim \sum_{\lambda \in \vee_\Psi} \|w_\lambda\|_V^2 < \infty$,
on $Y \times Y$, we define the bounded, symmetric, coercive bilinear form
\[
  (D_Y \sum_{\lambda \in \vee_{\Phi}} \psi_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_{\Phi}} \psi_\mu \otimes v_\mu ) := \sum_{\lambda \in \vee_\Psi} \la w_\lambda, v_\lambda \ra_V.
\]
With $E_Y^\delta$ the trivial embedding $Y^\delta \to Y$,
we set $D_Y^\delta := {E_Y^\delta}' D_Y E_Y^\delta \in \Lis(Y^\delta, {Y^\delta}')$,
whose norm and norm of its inverse are uniformly bounded in $\delta \in \Delta$.
Equipping $Y^\delta$ with a basis of the form $\cup_\lambda \psi_\lambda \otimes \Xi_\lambda^\delta$,
the matrix representation of $D_Y^\delta$ reads
\[
  \mathbf D_Y^\delta = \blockdiag [\mathbf A_\lambda^\delta]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Xi_\lambda^\delta, \Xi_\lambda^\delta \ra_V.
\]
For given $\mathbf B_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta)^{-1}$, we have
\[
  (\mathbf D_Y^\delta)^{-1} \eqsim \blockdiag[\mathbf B_\lambda^\delta]_\lambda.
\]
Suitable $\mathbf B_\lambda^\delta$, that can also be applied in linear complexity, are provided by \emph{multi-level methods},
e.g.~one can use a V-cycle method based on the stable subspace decomposition introduced in~\cite{Wu2017}.
 A simpler approach, although at superlinear complexity, would be to perform direct
solves in the spatial axis, i.e.~$\mathbf B_\lambda^\delta := (\mathbf A_\lambda^\delta)^{-1}$.

\subsubsection{A preconditioner for bilinear forms on $X \times X$}
Let $\Sigma$ be a Riesz basis for $L_2(I)$ such that
$\{2^{-|\lambda|}\sigma_\lambda : \lambda \in \vee_\Sigma\}$ is a Riesz basis for $H^1(I)$.
Using that any $w \in X$ is of the form $\sum_{\lambda \in \vee_\Sigma} \sigma_\lambda \otimes w_\lambda$ for some $w_\lambda \in V$,
on $X \times X$ we define the bilinear form
\[
  (D_X \sum_{\lambda \in \vee_{\Sigma}} \sigma_\lambda \otimes w_\lambda)(\sum_{\mu \in \vee_{\Sigma}} \sigma_\mu \otimes v_\mu) := \sum_{\lambda \in \vee_\Sigma} \la w_\lambda, v_\lambda \ra_V + 4^{|\lambda|} \la w_\lambda, v_\lambda \ra_{V'}.
\]
\begin{lemma}
  The symmetric bilinear form $D_X$ is bounded and coercive.
\end{lemma}
\begin{proof}
  The result follows directly from the norm equivalence
  \[
    \|w\|_X^2 \eqsim \sum_{\lambda \in \vee_\Sigma} \|w_\lambda\|_V^2 + 4^{|\lambda|} \|w_\lambda\|_{V'}^2.
  \]
  \jw{TODO prove; hier kwamen we vorige keer niet netjes uit.}
\end{proof}
Denoting with $E_X^\delta$ the trivial embedding $X^\delta \to X$, we again set
$D_X^\delta := {E_X^\delta}' D_X E_X^\delta \in \Lis(X^\delta, {X^\delta}')$, whose
norm and norm of its inverse are bounded uniformly in $\delta \in \Delta$. Equipping
$X^\delta$ with a basis of the form $\cup_\lambda \sigma_\lambda \otimes \Phi_\lambda^\delta$,
the matrix representation of $D_X^\delta$ reads
\[
  \mathbf D_X^\delta = \blockdiag[\mathbf A_\lambda^\delta + 4^{|\lambda|} \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'}]_\lambda, \quad \text{where} \quad \mathbf A_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_V.
\]
\begin{theorem}
If the collection $\mathcal O := \{W_\lambda^\delta = \tilde W_\lambda^\delta : \lambda \in \vee_\Sigma^\delta\}$
satisfies the conditions of Proposition~\ref{prop:space-infsup} with uniform inf-sup constant $\gamma>0$, and one has $\mathbf C_\lambda^\delta$ such that
\begin{equation}\label{eqn:robustmg}
    \mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1} \quad \text{uniformly in $\delta$ and $|\lambda|\in \N_0$},
\end{equation}
where $M_\lambda^\delta := \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra$, then
\[
    (\mathbf D_X^\delta)^{-1} \eqsim \blockdiag[\mathbf C_\lambda^\delta \mathbf A_\lambda^\delta \mathbf C_\lambda^\delta]_\lambda.
\]
  In other words, constructing the $\mathbf C_\lambda^\delta$ leads to an optimal preconditioner on $X \times X$.
\end{theorem}
\begin{proof}

  With $\mathbf u$ the representation of $u$ in $\Phi_\lambda^\delta$, we have
  $\|u\|_{{Y^\delta}'} = \|(\mathbf A_\lambda^\delta)^{-1/2} \mathbf M_\lambda^\delta \mathbf u\|$.
For $u \in W_\lambda^\delta$, we have $\gamma \|u\|_{V'} \leq \|u\|_{(W_\lambda^\delta)'} \leq \|u\|_{V'}$, so that
  \[
    \gamma^2 \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'} \leq \mathbf M_\lambda^\delta (\mathbf A_\lambda^\delta)^{-1} \mathbf M_\lambda^\delta \leq \la \Phi_\lambda^\delta, \Phi_\lambda^\delta \ra_{V'}.
  \]
  From \cite[Thm.~4]{Pearson2012}, we find the equivalence
  \[
\tfrac12\big( {\bf A}_\lambda^\delta + 4^{|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda\big)
\leq ({\bf A}_\lambda^\delta + 2^{|\lambda|}{\bf M}^\delta_\lambda) ({\bf A}_\lambda^\delta)^{-1} ({\bf A}_\lambda^\delta + 2^{|\lambda|}{\bf M}^\delta_\lambda)
\leq {\bf A}_\lambda^\delta + 4^{|\lambda|} {\bf M}^\delta_\lambda ({\bf A}^\delta_\lambda)^{-1}  {\bf M}^\delta_\lambda.
\]
Therefore, given $\mathbf C_\lambda^\delta \eqsim (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$ uniformly in $\delta$ and $|\lambda|$, the result holds.
\end{proof}
In~\cite{olshanskii2000convergence, kolev2016multilevel} it was shown that a V-cycle multi-grid method yields a ${\bf C}^\delta_{\lambda}$ that satisfies~\eqref{eqn:robustmg}, and that moreover  can be applied in linear complexity.
These results were shown for uniformly refined meshes, but similarly as for $4^{-|\lambda|} = 0$, we expect them to hold for locally refined meshes as well. \rvv{Hier had Rob iets snuggers over te zeggen.}

 A (suboptimal) alternative is to take
$\mathbf C_\lambda^\delta := (\mathbf A_\lambda^\delta + 2^{|\lambda|} \mathbf M_\lambda^\delta)^{-1}$
directly.
